---
title: "GB3 Kinetic Ensemble Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GB3 Kinetic Ensemble Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Datasets

## 2LUM Three-State Ensemble

The [2LUM](https://www.rcsb.org/structure/2LUM) 60-member GB3 ensemble was generated by joint refinement of three-state ensembles to eNOE, scalar coupling, and RDC data as described in [Vögeli 2012](https://doi.org/10.1038/nsmb.2355). There are two versions of the 2LUM ensemble included with the `ke` package. One contains all C-alpha atoms of the 60-member ensemble. The other contains all atoms of just models 1, 2, 21, 22, 41, and 42. They are read into three-dimensional coordinate arrays, with the first dimension representing x/y/z, the second representing atoms, and the third representing the models:

```{r}
library(ke)

ca_file <- system.file("extdata", "gb3", "2lum_ca.pdb.gz", package = "ke")
ca_coord <- read_ensemble(ca_file)
str(ca_coord)
subset_file <- system.file("extdata", "gb3", "2lum_subset.pdb.gz", package = "ke")
subset_coord <- read_ensemble(subset_file)
str(subset_coord)
subset_models <- c(1, 2, 21, 22, 41, 42)
```

The following carries out a principal components analysis on the C-alpha atoms and projects them onto the first two principal components.

```{r}
ca_matrix <- t(apply(ca_coord, 3, c))
ca_matrix <- scale(ca_matrix, center = TRUE, scale = FALSE)

ca_pca <- eigen(cov(ca_matrix))
ca_proj <- ca_matrix %*% ca_pca$vectors[, 1:2]
```

The first two principal component projections can be plotted, revealing reasonably distinct states.

```{r, fig.width = 5, fig.height = 5}
model_col <- c(rep("black", 20), rep("red", 20), rep("blue", 20))
plot(ca_proj[, 1], ca_proj[, 2], xlab = "PC1 projection", ylab = "PC2 projection",
     main = "2LUM C-alpha PCA", type = "n")
text(ca_proj[, 1], ca_proj[, 2], labels = 1:60, col = model_col)
points(ca_proj[subset_models, 1], ca_proj[subset_models, 2], 
       pch = 1, col = "purple", cex = 3, lwd = 2)
legend("topright", legend = c("Models 1-20", "Models 21-40", "Models 41-60"), 
       text.col = c("black", "red", "blue"), bty = "n")
legend("bottomright", legend = "Subset models", col = "purple", 
       pch = 1, pt.cex = 3, pt.lwd = 2, bty = "n")
```

The six models included in `2lum_subset.pdb.gz` are circled in purple above. Working under the assumption that models refined together have numbers separated by 20, those should be the first two three-state ensembles.

## <sup>15</sup>N Backbone Relaxation Data

The `hall2006` directory contains GB3 <sup>15</sup>N backbone relaxation data reported by [Hall 2006](https://doi.org/10.1021/ja060406x): longitudinal relaxation rates (`R1.csv`), transverse relaxation rates (`R2.csv`), and steady-state heteronuclear NOE values (`NOE.csv`). Each file contains one row per residue, with measured values and associated uncertainties at five proton spectrometer frequencies ranging from approximately 400 to 800 MHz (9.4-18.8 T, respectively).

```{r}
hall2006_dir <- system.file("extdata", "gb3", "hall2006", package = "ke")

r1 <- read.csv(file.path(hall2006_dir, "R1.csv"))
r2 <- read.csv(file.path(hall2006_dir, "R2.csv"))
noe <- read.csv(file.path(hall2006_dir, "NOE.csv"))

str(r1)
str(r2)
str(noe)
```

The steady-state heteronuclear NOE values can be combined with the
corresponding `R1` values to estimate sigma cross-relaxation rates. For
example, the following code constructs a sigma data frame in the same wide
format as the `R1`, `R2`, and `NOE` tables.

```{r}
field_labels <- c("9.4T", "11.7T", "14.1T", "16.4T", "18.8T")

sigma <- data.frame(residue = r1$residue)

for (field in field_labels) {
  sigma_field <- noe_to_sigma(
    noe = noe[[paste0("NOE_", field)]],
    r1 = r1[[paste0("R1_", field)]],
    dnoe = noe[[paste0("NOE_err_", field)]],
    dr1 = r1[[paste0("R1_err_", field)]]
  )
  sigma[[paste0("sigma_", field)]] <- sigma_field[, "sigma"]
  sigma[[paste0("sigma_err_", field)]] <- sigma_field[, "sigma_err"]
}

str(sigma)
```

# Input Data

## <sup>15</sup>N Backbone Relaxation Data

To construct relaxation input for the subset ensemble, the Hall 2006 residue
tables can be matched to backbone amide N-H pairs in `subset_coord`. The
first step is to identify which backbone amide N-H pairs are present in the
subset ensemble and keep only those residue numbers that overlap the Hall 2006
relaxation tables.

```{r}
atom_ids <- dimnames(subset_coord)[[2]]
atom_meta <- strcapture(
  "^\\s*([A-Z0-9]+)\\s+([A-Z]{3})\\s+A\\s+([0-9]+)\\s*$",
  atom_ids,
  proto = list(atom_name = character(), res_name = character(), residue = integer())
)
atom_meta$atom_id <- atom_ids

n_atom <- atom_meta[atom_meta$atom_name == "N", c("residue", "atom_id")]
h_atom <- atom_meta[atom_meta$atom_name == "H", c("residue", "atom_id")]
names(n_atom)[2] <- "atom1"
names(h_atom)[2] <- "atom2"

atom_pairs <- merge(n_atom, h_atom, by = "residue")
atom_pairs <- atom_pairs[order(atom_pairs$residue), ]

relax_table <- Reduce(
  function(x, y) merge(x, y, by = "residue"),
  list(r1, r2, sigma)
)

relax_table <- merge(atom_pairs, relax_table, by = "residue")
atom_pairs <- relax_table[, c("atom1", "atom2")]

str(atom_pairs)
```

Next, the experimental `R1`, `R2`, and sigma values are converted into the
`relax_data_list` format used by the package. Each entry in the list contains
both a numeric vector of observed values and a spectral-density term array
describing how that rate should be calculated for a given field strength. The
following code constructs a 15-element list representing three types of
relaxation rates (R<sub>1</sub>, R<sub>2</sub>, and $\sigma_\mathrm{NH}$) at
each of five field strengths.

```{r}

field_mhz <- c("9.4T" = 400, "11.7T" = 500, "14.1T" = 600, "16.4T" = 700, "18.8T" = 800)
r_nh_angstrom <- 1.02
delta_sigma_ppm <- -172
k_from_err <- function(err) {
  err_floor <- min(err[is.finite(err) & err > 0], na.rm = TRUE)
  1 / pmax(err, err_floor)^2
}

relax_data_list <- list()

for (field in names(field_mhz)) {
  proton_mhz <- field_mhz[[field]]
  relax_data_list[[paste0("r1_", proton_mhz)]] <- list(
    value = relax_table[[paste0("R1_", field)]],
    k = k_from_err(relax_table[[paste0("R1_err_", field)]]),
    spec_den_term_array = make_r1_spec_den_term_array(
      n_pairs = nrow(atom_pairs),
      proton_mhz = proton_mhz,
      r_is_angstrom = r_nh_angstrom,
      delta_sigma_ppm = delta_sigma_ppm
    )
  )
  relax_data_list[[paste0("r2_", proton_mhz)]] <- list(
    value = relax_table[[paste0("R2_", field)]],
    k = k_from_err(relax_table[[paste0("R2_err_", field)]]),
    spec_den_term_array = make_r2_spec_den_term_array(
      n_pairs = nrow(atom_pairs),
      proton_mhz = proton_mhz,
      r_is_angstrom = r_nh_angstrom,
      delta_sigma_ppm = delta_sigma_ppm
    )
  )
  relax_data_list[[paste0("sigma_", proton_mhz)]] <- list(
    value = relax_table[[paste0("sigma_", field)]],
    k = k_from_err(relax_table[[paste0("sigma_err_", field)]]),
    spec_den_term_array = make_sigma_spec_den_term_array(
      n_pairs = nrow(atom_pairs),
      proton_mhz = proton_mhz,
      nucleus_i = "1H",
      nucleus_s = "15N",
      r_is_angstrom = r_nh_angstrom
      )
  )
}

str(relax_data_list, max.level = 1)
```

It can be helpful to inspect one rate block in more detail. In the example
below, `r1_400` contains the observed *R*<sub>1</sub> values at 400 MHz
and the corresponding spectral-density coefficients and frequencies used to
calculate those rates. The `value` vector gives the measured
*R*<sub>1</sub> values.

```{r}
str(relax_data_list[["r1_400"]])
```

The first `spec_den_term_array` array dimension matches the rows of
`atom_pairs`. A small residue subset can be used to see exactly how the
values and term arrays line up.

```{r}
example_idx <- 1:4
data.frame(
  residue = relax_table$residue[example_idx],
  atom_pairs[example_idx, ],
  r1_400 = relax_data_list[["r1_400"]]$value[example_idx]
)

relax_data_list[["r1_400"]]$spec_den_term_array[example_idx, , ]
```

For backbone <sup>15</sup>N relaxation, the *R*<sub>1</sub> expression below
has been rewritten by spectral-density term to match the structure of
`spec_den_term_array`. Here, *d* and *c* represent the dipolar and chemical
shift anisotropy constants, respectively.

$$
R_1 =
\left(\frac{1}{10} d_{NH}^2\right) J(\omega_H - \omega_N) +
\left(\frac{3}{10} d_{NH}^2 + c_N^2\right) J(\omega_N) +
\left(\frac{3}{5} d_{NH}^2\right) J(\omega_H + \omega_N).
$$
The names used in the second dimension of the spectral-density term arrays
follow a short frequency-label convention. In the case of *R*<sub>1</sub>,

- `wHmwN` means $\omega_H - \omega_N$
- `wN` means $\omega_N$
- `wHpwN` means $\omega_H + \omega_N$

In other equations,

- `2wH` means $2 \, \omega_H$
- similarly, `wC`, `wF`, `wP`, and `wD` denote the corresponding nucleus
  frequencies

The third dimension stores the `coef` and `freq` components for each term. In
other words, the `value` vector gives the measured *R*<sub>1</sub> values, and
`spec_den_term_array` gives the weighted $J(\omega)$ terms used to calculate
those same relaxation rates.

Finally, the atom pairs, the `unit` flag, and the relaxation
blocks can be assembled into a single `atom_relax_data` object. This is the
input structure used to build the kinetic-model
matrices needed for relaxation calculations.

```{r}
atom_relax_data <- list(
  atom_pairs = atom_pairs,
  unit = TRUE,
  relax_data_list = relax_data_list
)

str(atom_relax_data, max.level = 2)
```

To evaluate these relaxation expressions for a kinetic ensemble, the next step
is to specify a kinetic model for exchange among the ensemble members. As a
simple example, the code below assumes that all six members of
`subset_coord` exchange through a single effective rate constant,
`kens`. This defines a transition-rate matrix and then uses
`make_spec_den_relax_data()` to combine the kinetic model with the
`atom_relax_data` structure prepared above.

```{r}
base_rates <- c(kens = 2e9)
subset_coord <- subset_coord[, , 1:6, drop = FALSE]
base_rate_mat <- rate_mat_simple(base_rates, dimnames(subset_coord)[[3]])

spec_den_relax_data <- make_spec_den_relax_data(
  atom_relax_data = atom_relax_data,
  base_rate_mat = base_rate_mat,
  base_rates = base_rates
)

str(spec_den_relax_data, max.level = 2)
```

# Writing Spectral Density Relaxation Data

The `spec_den_relax_data` object can also be written to disk as a set of CSV
files. This is useful when preparing an analysis outside the current R session,
or when inspecting the exact inputs that [read_spec_den_relax_data()] expects to
reconstruct.

```{r}
relax_prefix <- file.path(tempdir(), "gb3_15n")
write_spec_den_relax_data(spec_den_relax_data, relax_prefix)

relax_files <- list.files(
  tempdir(),
  pattern = "^gb3_15n_(atom_relax|groupings|a_coef|lambda_coef)\\.csv$",
  full.names = TRUE
)

basename(relax_files)
```

The `*_atom_relax.csv` file stores the atom-pair identifiers together with the
flattened relaxation-rate definitions. Each rate contributes optional `value`
and `k` columns, followed by one `coef` and one `freq` column for each
spectral-density term. To keep the example compact, the table below shows the
first relaxation-rate block, `r1_400`, for a small subset of residues.

```{r}
atom_relax_csv <- read.csv(paste0(relax_prefix, "_atom_relax.csv"), check.names = FALSE)
first_rate <- names(spec_den_relax_data$relax_data_list)[[1]]
first_rate_cols <- grep(paste0("^", first_rate, "_"), names(atom_relax_csv), value = TRUE)

atom_relax_subset <- atom_relax_csv[
  seq_len(6),
  c(names(atom_relax_csv)[1:2], first_rate_cols),
  drop = FALSE
]

knitr::kable(atom_relax_subset[, 1:6, drop = FALSE], digits = 3)
knitr::kable(atom_relax_subset[, 7:10, drop = FALSE], digits = 3)
```

The `*_groupings.csv` file encodes how internal kinetic modes are grouped for
dipole-dipole interaction tensor averaging. Each row corresponds to one
grouping pattern used later when constructing the relaxation calculation.

```{r}
groupings_csv <- as.matrix(
  read.csv(paste0(relax_prefix, "_groupings.csv"), header = FALSE)
)

knitr::kable(groupings_csv, col.names = NULL)
```

The `*_a_coef.csv` file stores the coefficients used to build the internal-mode
amplitudes from the supplied kinetic model. Its columns correspond to the base
rate constants, and its rows correspond to grouped internal amplitudes.

```{r}
a_coef_csv <- read.csv(paste0(relax_prefix, "_a_coef.csv"), check.names = FALSE)

knitr::kable(a_coef_csv, digits = 3)
```

Finally, the `*_lambda_coef.csv` file stores the coefficients used to assemble
the internal decay rates. These coefficients are combined with the named base
rates to produce the internal eigenvalues used in the spectral-density sums.

```{r}
lambda_coef_csv <- read.csv(
  paste0(relax_prefix, "_lambda_coef.csv"),
  check.names = FALSE,
  row.names = 1
)

knitr::kable(lambda_coef_csv, digits = 3)
```

# Fiting *k*<sub>ens</sub> and *D*<sub>iso</sub>

Because `atom_relax_data` stores one force constant per relaxation rate, the
same setup can also be used in a simple numerical fit. The example below uses
[optim()] to refine the ensemble exchange rate `kens` and an isotropic
rotational diffusion constant `Diso` by minimizing the weighted quadratic loss
returned by [coord_array_to_relax_energy()]. The optimization is carried out on
the log scale so that both fitted parameters remain positive.

```{r}
coord_array_opt <- aperm(subset_coord, c(2, 1, 3))
quadratic_loss_local <- function(x, x0, k = 1, gradient = FALSE) {
  expr1 <- x - x0
  value <- k * expr1^2
  if (gradient) {
    attr(value, "gradient") <- k * (2 * expr1)
  }
  value
}

relax_energy_objective <- function(log_par) {
  kens <- exp(log_par[[1]])
  diso <- exp(log_par[[2]])
  rates_fit <- c(kens = kens, Dx = diso, Dy = diso, Dz = diso)

  coord_array_to_relax_energy(
    coord_array = coord_array_opt,
    rates = rates_fit,
    spec_den_relax_data_list = list("15N" = spec_den_relax_data),
    loss_func = quadratic_loss_local
  )
}

optim_fit <- optim(
  par = log(c(kens = 2e9, Diso = 5e7)),
  fn = relax_energy_objective,
  method = "Nelder-Mead"
)

fit_rates <- c(
  kens = exp(optim_fit$par[[1]]),
  Dx = exp(optim_fit$par[[2]]),
  Dy = exp(optim_fit$par[[2]]),
  Dz = exp(optim_fit$par[[2]])
)
fit_summary <- c(
  kens = fit_rates[["kens"]],
  tau_ens_ns = 1e9 / fit_rates[["kens"]],
  Diso = fit_rates[["Dx"]],
  tau_c_ns = 1e9 / (6 * fit_rates[["Dx"]])
)

fit_summary
```

# Plotting Relaxation Rates

The optimized values can then be inserted into the named rate vector and used
with [coord_array_to_relax()] to calculate relaxation rates from the ensemble
coordinates.

```{r}
relax_calc <- coord_array_to_relax(
  coord_array = coord_array_opt,
  rates = fit_rates,
  spec_den_relax_data_list = list("15N" = spec_den_relax_data)
)[["15N"]]

str(relax_calc)
```

The plots below compare the experimental data to the rates calculated from the
ensemble. Each color corresponds to one field strength, the closed circles show
experimental values, the vertical bars show experimental uncertainties, and
the open circles show corresponding rates calculated from the ensemble.

```{r, fig.width = 7, fig.height = 10}
field_colors <- c(
  "400" = "#1b9e77",
  "500" = "#d95f02",
  "600" = "#7570b3",
  "700" = "#e7298a",
  "800" = "#66a61e"
)
field_tesla_labels <- setNames(names(field_mhz), as.character(field_mhz))

plot_relax_comparison <- function(rate_prefix, value_prefix, err_prefix, ylab) {
  residue <- relax_table$residue
  calc_cols <- grep(paste0("^", rate_prefix, "_"), colnames(relax_calc), value = TRUE)
  field_names <- sub(paste0("^", rate_prefix, "_"), "", calc_cols)
  field_labels_t <- unname(field_tesla_labels[field_names])
  xlim <- c(min(residue), max(residue) + 6)

  y_exp <- unlist(lapply(field_labels_t, function(field) relax_table[[paste0(value_prefix, field)]]))
  y_err <- unlist(lapply(field_labels_t, function(field) relax_table[[paste0(err_prefix, field)]]))
  y_calc <- unlist(lapply(calc_cols, function(col) relax_calc[, col]))

  plot(
    xlim, range(c(y_exp - y_err, y_exp + y_err, y_calc), na.rm = TRUE),
    type = "n", xlab = "Residue", ylab = ylab
  )

  for (i in seq_along(field_names)) {
    field <- field_names[[i]]
    field_t <- field_labels_t[[i]]
    col <- field_colors[[field]]
    exp_y <- relax_table[[paste0(value_prefix, field_t)]]
    err_y <- relax_table[[paste0(err_prefix, field_t)]]
    calc_y <- relax_calc[, paste0(rate_prefix, "_", field)]
    has_err <- err_y > 0 & !is.na(err_y)

    graphics::arrows(
      x0 = residue[has_err], y0 = exp_y[has_err] - err_y[has_err],
      x1 = residue[has_err], y1 = exp_y[has_err] + err_y[has_err],
      angle = 90, code = 3, length = 0.03, col = col
    )
    points(residue, exp_y, pch = 16, col = col)
    points(residue, calc_y, pch = 1, col = col)
  }

  legend(
    "right",
    legend = paste0(field_names, " MHz"),
    col = unname(field_colors[field_names]),
    pch = 16,
    bty = "n"
  )
}

old_par <- par(mfrow = c(3, 1), mar = c(4, 4, 2, 4))
plot_relax_comparison("r1", "R1_", "R1_err_", expression(R[1]))
plot_relax_comparison("r2", "R2_", "R2_err_", expression(R[2]))
plot_relax_comparison("sigma", "sigma_", "sigma_err_", expression(sigma[NH]))
par(old_par)
```

The following grid shows the same comparison broken out by both relaxation-rate
type and field strength. Each panel plots the modeled rates against the
experimental values for one field, with vertical error bars showing the
experimental uncertainties and the dashed diagonal marking perfect agreement.

```{r, fig.width = 7, fig.height = 5}
plot_modeled_vs_experimental <- function(rate_prefix, value_prefix, err_prefix, main_prefix) {
  field_names <- names(field_mhz)

  for (field in field_names) {
    proton_mhz <- field_mhz[[field]]
    exp_y <- relax_table[[paste0(value_prefix, field)]]
    err_y <- relax_table[[paste0(err_prefix, field)]]
    calc_y <- relax_calc[, paste0(rate_prefix, "_", proton_mhz)]
    keep <- is.finite(exp_y) & is.finite(calc_y)
    lim <- range(c(exp_y[keep] - err_y[keep], exp_y[keep] + err_y[keep], calc_y[keep]), na.rm = TRUE)

    plot(
      exp_y, calc_y,
      xlab = "Experimental",
      ylab = "Modeled",
      main = paste(main_prefix, proton_mhz, "MHz"),
      pch = 16,
      col = field_colors[[as.character(proton_mhz)]],
      xlim = lim,
      ylim = lim
    )
    abline(0, 1, lty = 2, col = "gray50")
    has_err <- keep & err_y > 0 & !is.na(err_y)
    graphics::arrows(
      x0 = exp_y[has_err] - err_y[has_err], y0 = calc_y[has_err],
      x1 = exp_y[has_err] + err_y[has_err], y1 = calc_y[has_err],
      angle = 90, code = 3, length = 0.03,
      col = field_colors[[as.character(proton_mhz)]]
    )
  }
}

old_par <- par(mfrow = c(3, 5), mar = c(3, 3, 2, 1), mgp = c(1.7, 0.6, 0))
plot_modeled_vs_experimental("r1", "R1_", "R1_err_", expression(R[1]))
plot_modeled_vs_experimental("r2", "R2_", "R2_err_", expression(R[2]))
plot_modeled_vs_experimental("sigma", "sigma_", "sigma_err_", expression(sigma))
par(old_par)
```
